Towards a Verified Bzip2: A Functional Specification for Burrows--Wheeler

### 2019-05-06 draft_paper/draft_paper.tex
Stash

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows--Wheeler \cite{appel-func-spec}. But more
generally, this provides evidence that Bird's approach of program
calculation, in which programs are derived from their specifications
via algebraic manipulations, is useful for proving the correctness of
algorithms in a machine-checked setting. Bird has applied this
approach to greedy algorithms, the Boyer-Moore algorithm, the
Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding, and
the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
more problems \cite{pearls}.

As a part of the proof of \Cref{thm:unbwt_correct}, we proved two
other noteworthy results that have not, to our knowledge, been proved
before in Coq. We implemented a least-significant-digit (LSD)
radixsort and proved its correctness,
\begin{theorem}[radixsort\_correct]
  \label{thm:radixsort_correct}%
  For all matrices $m$ with $n$ columns,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~m~n) \land \var{Permutation}~(\var{radixsort}~m~n)~m
  \end{equation*}
\end{theorem}
and we proved that any two stable sorts are interchangable,
\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,~f~l = g~l)
  \end{equation*}
\end{restatable*}
This in turn required developing suitable definitions for
\var{Sorted}, \var{Permutation}, and \var{StablePerm}. We explored
multiple definitions for each of these properties, and proved
equivalence between them.

@@

\section{History and Applications of the BWT}
\label{sec:bwt}

@@

The next pass is a Move-to-Front (MTF) encoding, in which a character
is represented by the number of different characters encountered since
the last occurence of this character. This creates many runs of zeroes
from the runs in the output of BWT. On English texts, often 50-60\% of
the values produced by the MTF pass are zeroes \cite{fenwick2007}.
Often a Run-Length Encoding (RLE) pass is then used to encode only the
runs of zeroes with their lengths, using a bijective base-2 encoding
\cite{bw-analysis, tsai_2016}. Finally, some sort of zeroth-order
coder like a Huffman coder is used.

@@

\section{Orders and Preorders}
\label{sec:ord}

@@

The builtin class \var{EqDec} is used to describe types with decidable
equivalence relations, so for any \var{Preord} instance we have the
\var{EqDec} instance \var{Preord\_EqDec}. In Coq, the notation
\var{===} is used for the equivalence relation of an \var{EqDec}
instance and \var{==} for its decision procedure.

@@

\section{The Forwards BWT}
\label{sec:forwards_BWT}

@@

Now we would likel to produce the rotation matrix by repeatedly
applying \var{lrot}. We will use \var{iter}, which takes a function
$f : A \to A$, a number $n$, and an initial value $z$, and returns the
list of intermediate values obtained by applying $f$ to $z$ $n$ times.
Or,
\begin{equation}
  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^n z]
  \label{eq:iter}
\end{equation}
\lstinputlisting[
  widthgobble=2,
  firstline=14,
  lastline=18
]{../theories/BWT/Lib/Iterate.v}

We can prove that this definition satisfies \Cref{eq:iter} by defining
the function \var{rep}\footnote{Alternatively, we can say that
  $\lambda f.~\var{rep}~n~f$ is the $n$th Church numeral.} for
repeated function application.
\lstinputlisting[
  widthgobble=2,
  firstline=8,
  lastline=12
]{../theories/BWT/Lib/Repeat.v}
In proofs, we will still use the notation $f^i z$ for $\var{rep}~f~i~z$.

@@

\var{bwl} produces the transformed string by taking the last column
of the sorted rotation matrix. \var{lexsort} sorts the rotation
matrix lexicographically using the \var{Ord} instance; we will
discuss it's implementation in \cref{sec:sort}.
\lstinputlisting[
  widthgobble=2,
  firstline=31,
  lastline=35
]{../theories/BWT/BurrowsWheeler.v}

@@

\section{The Inverse BWT}
\label{sec:inverse_BWT}

@@

We also can notice that the columns to the right of the sorted column
move with it for similar reasons. The column directly to its right is
the result of sorting the input string, and then stably sorting it
again by the character that precedes it in the original string. The
next column is the result of sorting by each character, sorting by the
preceding chracter, and then sorting by the character 2 spaces in
front.

All this means that the columns to the left of the sorted column seem
like they don't matter, which suggests that we can recreate the entire
rotation matrix from just the last column (\Cref{fig:recreate}).
However, the column direclty to the left of the sorted column is very
important, as it is the next column to be sorted. And we can see that
this column changes, only becoming the eventual last column of the
matrix in the second to last iteration. Why then, does it still work
if we use the last column everywhere?

We can think of the process of sorting the matrix as computing the
last column subject to the constraints of the lexicographic ordering.
For example, we know that the collection of all characters preceding
an \texttt{a} in the sorted column must be some permutation of
$[\texttt{c},~\texttt{d},~\texttt{r},~\texttt{r},~\texttt{!}]$,
because those are the characters that preced \texttt{a}s in the
original string. However, the order of these characters in the final
string
($[\texttt{r},~\texttt{d},~\texttt{!},~\texttt{r},~\texttt{c}]$) is
important. An \texttt{r} must be first, because in the original string
there is an \texttt{a} following an \texttt{r} which is itself
followed by a \texttt{!}, the smallest character. Similarly, although
the \texttt{d} and \texttt{!} precede \texttt{a}s that are both
followed by \texttt{bra}s, the next character for the \texttt{d} is a
\texttt{!}, and the next character for the \texttt{!} is a \texttt{c}.
This requires that the \texttt{d} preced the \texttt{!} in the last
column.

@@

The word \textit{permutation} has two related meanings that are often
used interchangably. We sometimes say that two lists are permutations of each
other if they contain all of the same elements, just appearing in a
different order. However, in mathematics we often say that a
permutation is actually the bijection on indices that defines the
correspondence between the two lists. For example, we say that $p$ is
the permutation that relates $l$ to $l'$ if for all indices $i$ and
$j$,
\begin{equation*}
  l_{p(i)} = l'_i
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
each other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
permutations of each other is precisely when
\begin{itemize}
  \item $\var{length}~l = \var{length}~l' = n$
  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
    \begin{equation*}
      l_{p(i)} = l'_i
    \end{equation*}
\end{itemize}

@@

To restrict the range to $\mathbb{N}_{<n}$ we say that $\forall i,~i \in p
\implies ~i<n$, to make the function surjective, we strengthen that to
$\forall i,~i \in p \iff ~i<n$, and to make it injective we require that $p$ not
contain any duplicate entries. The inductive property \var{NoDup} from
the Coq standard library captures that requirement:
\begin{align*}
  \infer[\var{NoDup\_nil}]{\var{NoDup}~[]}{}
  &&
  \infer[\var{NoDup\_cons}]{\var{NoDup}~(x :: l)}{%
    x \notin l & \var{NoDup}~l
  }
\end{align*}

@@

\begin{theorem}[apply\_correct]
  For all permutation functions $p$ and lists $l$,
  \begin{equation*}
    \var{Permutation}~(\var{apply}~p~l)~l
  \end{equation*}
\end{theorem}
\begin{proof}
  First we let $n = \var{length}~l$, so that we can reason by
  induction on $n$. The base case is uninteresting. In the inductive
  case, we have the induction hypothesis that for all lists of length
  $n$ and permutation functions on such lists,
  \begin{equation*}
    \var{Permutation}~(\var{apply}~p~l)~l
  \end{equation*}
  and we must show that
  \begin{equation*}
    \var{Permutation}~(\var{apply}~(i :: p)~(a :: l))~(a :: l)
  \end{equation*}

  We reason as follows, using the notation $\eqperm$ to
  chain together \var{Permutation}s:
  \begin{align*}
    \var{apply}~(i :: p')~(a :: l) &\eqperm
    (a :: l)_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a::l)) \\
    &\eqperm (a :: l)_i :: \var{rem\_nth}~i~(a::l) \\
    &\eqperm a :: l
  \end{align*}
  where the first step follows from \var{rem\_PermFun\_correct}, the
  second from the induction hypothesis, and the last from
  \var{rem\_nth\_Perm}.
\end{proof}

@@

For completeness, we will briefly mention a third definition of a
permutation relation that says that two lists are permutations of one
another when any element occurs an equal number of times in each:
\lstinputlisting[
  widthgobble=2, firstline=14, lastline=15,
]{../theories/BWT/Sorting/PermutationCount.v}
\var{count\_occ} counts the number of occurences of a given variable
using a decision procedure for Leibniz equality. Here that decision
procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
require the equivalence relation itself by Leibniz equality.

@@

Based on this, we define a stronger version of the \var{PermFun}
property, \var{StablePermFun}. However, while $\var{PermFun}~n~p$
means that $p$ represents a permutation function on lists of length
$n$, we need to specify the list with resepct to which $p$ is stable,
so we say $\var{StablePermFun}~l~p$ for some list $l$.
\lstinputlisting[
  widthgobble=2, firstline=557, lastline=561,
]{../theories/BWT/Sorting/PermFun.v}
As above, we can also define this property using the inverse
permutation:
\lstinputlisting[
  widthgobble=2, firstline=563, lastline=567,
]{../theories/BWT/Sorting/PermFun.v}
As usual, we will prove these definitions equivalent. First, though,
we need to define \var{image} and \var{preimage}.
\lstinputlisting[
  widthgobble=2, firstline=85, lastline=86,
]{../theories/BWT/Sorting/PermFun.v}
We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$
and are invereses of each other. We also have the following two
theorems:
\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
  For all lists $l$, permutations $p$, and indices $i$,
  \begin{align*}
    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
  \end{align*}
\end{theorem}
\begin{proof}
  The first equation (\var{nth\_preimage\_apply}) holds because:
  \begin{align*}
       (\var{apply}~p~l)_{p^{-1}(i)}
    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
    &= l_{p_{(\var{findIndex}~i~p)}} \\
    &= l_i
  \end{align*}
  and the second follows from this:
  \begin{equation*}
      (\var{apply}~p~l)_i
    = (\var{apply}~p~l)_{p^{-1}(p(i))}
    = l_{p(i)}
    \qedhere
  \end{equation*}
\end{proof}

@@

Unforunately, neither of these two definitions will be particularly
useful for proving the theorems we would like to prove with them. The
best definition for stability does not come by analogy with
\var{Permutation}. The intuition is that for any two lists that are
stable permutations of each other, if we filter each list by
equivalence with any element, the resulting lists should be identical.
This is because they must have the same elements, since the original
lists are permutations of one another, and those elements must be in
the same order, since the original lists are specifically stable
permutations of one another.
\lstinputlisting[
  widthgobble=2, firstline=21, lastline=22,
]{../theories/BWT/Sorting/StablePerm.v}

@@

\section{Uniqueness of Stable Sorts}
\label{sec:unique}

@@

  First, we can say that $h \equiv h'$: we know that $h \in h' :: t'$ and $h'
  \in h :: t$. If $h = h'$, then we are done, so assuming $h \neq h'$ we
  know $h \in t'$ and $h' \in t$. Because both $h::t$ and $h'::t'$ are
  sorted, we knnow that $h \le h'$ and $h' \le h$, implying $h \equiv h'$.

@@

\section{Lexicographic Orders and Keys}
\label{sec:lex_ord}

@@

First we want to prove that \var{radixsort} permutes the rows of the
matrix. We will start by proving the invariant that if we rotate the
last column to the front and sort $j$ times, the resulting matrix is a
permtuation of simply rotating the last column to the front $j$
times.
\begin{lemma}[radixsort\_perm\_inv]
  For all matrices $m$ and $j : \mathbb{N}$,
  \begin{equation*}
    \var{Permutation}~((\var{map}~\var{rrot})^j~m)~
                      ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
  \end{equation*}
\end{lemma}
\begin{proof}
  We reason by induction on $j$. In the inductive case,
  \begin{align*}
    (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m ={}&
    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
    \var{map}~\var{rrot})^j~m) \\
    \eqperm{}& \var{map}~\var{rrot}~((\var{hdsort} \comp
    \var{map}~\var{rrot})^j~m) \\
    \eqperm{}& \var{map}~\var{rrot}~((\var{map}~\var{rrot})^j~m) \\
    ={}& (\var{map}~\var{rrot})^{j+1}~m \qedhere
  \end{align*}
\end{proof}

@@

@@

\section{Insertion Sort}
\label{appendix:insertion_sort}

Insertion sort is a simple and elegant stable sort that is easy to
reason about. The theorems proved in \cref{sec:unique} allow us to
prove things using insertion sort that apply to any stable sort. For
example, our proof of correctness for radixsort relies on the fact
that it uses insertion sort as the innner sort, but
\var{StableSort\_unique} allows us to extend that proof to a radixsort
that uses any stable sort. We also use it as the implementation of
\var{lexsort} because it is convenient, but similarly
\var{Sort\_Ord\_unique} ensures that it can be replaced with any other
sort.

@@

\section{Permutation Composition}
\label{appendix:perm_comp}

@@

We will use the function \var{combine} defined in the Coq standared
library to annotate every element of $l$ with its index.
\begin{lstlisting}
  Fixpoint combine (l : list A) (l' : list B) : list (A*B) :=
    match l,l' with
    | x :: tl, y :: tl' => (x,y) :: (combine tl tl')
    | _, _ => nil
    end.
\end{lstlisting}

>>>>

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows--Wheeler \cite{appel-func-spec}. But more
generally, this provides evidence that Bird's approach of program
calculation, in which programs are derived from their specifications
via algebraic manipulations, is useful for proving the correctness of
algorithms in a machine-checked setting. Bird has applied this
approach to greedy algorithms, the Boyer-Moore algorithm, the
Knuth-Morris-Pratt algorithm, Sudoku solvers, arithmetic coding, the
Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many more
problems \cite{pearls}.

As a part of the proof of \Cref{thm:unbwt_correct}, we proved two
other noteworthy results that have not, to our knowledge, been proved
before in Coq. We implemented a least-significant-digit (LSD)
radixsort and proved its correctness,
\begin{theorem}[radixsort\_correct]
  \label{thm:radixsort_correct}%
  For all matrices $m$ with $n$ columns,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~m~n) \land \var{Permutation}~(\var{radixsort}~m~n)~m
  \end{equation*}
\end{theorem}
and we proved that any two stable sorts are interchangeable,
\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,~f~l = g~l)
  \end{equation*}
\end{restatable*}
This in turn required developing suitable definitions for
\var{Sorted}, \var{Permutation}, and \var{StablePerm}. We explored
multiple definitions for each of these properties, and proved
equivalence between them.

@@

\section{History and applications of the BWT}
\label{sec:bwt}

@@

The next pass is a Move-to-Front (MTF) encoding, in which a character
is represented by the number of different characters encountered since
the last occurrence of this character. This creates many runs of zeroes
from the runs in the output of BWT. On English texts, often 50-60\% of
the values produced by the MTF pass are zeroes \cite{fenwick2007}.
Often a Run-Length Encoding (RLE) pass is then used to encode only the
runs of zeroes with their lengths, using a bijective base-2 encoding
\cite{bw-analysis, tsai_2016}. Finally, some sort of zeroth-order
coder like a Huffman coder is used.

@@

\section{Orders and preorders}
\label{sec:ord}

@@

The builtin class \var{EqDec} is used to describe types with decidable
equivalence relations, so for any \var{Preord} instance we have the
\var{EqDec} instance \var{Preord\_EqDec}. In Coq, the notation
\var{=\hspace{0pt}=\hspace{0pt}=} is used for the equivalence relation of an \var{EqDec}
instance and \var{=\hspace{0pt}=} for its decision procedure.

@@

\section{The forwards BWT}
\label{sec:forwards_BWT}

@@

Now we produce the rotation matrix by repeatedly applying \var{lrot}.
We will use \var{iter}, which takes a function $f : A \to A$, a number
$n$, and an initial value $z$, and returns the list of intermediate
values obtained by applying $f$ to $z$ $n$ times. Or,
\begin{equation}
  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^{n-1} z]
  \label{eq:iter}
\end{equation}
\lstinputlisting[
  widthgobble=2,
  firstline=14,
  lastline=18
]{../theories/BWT/Lib/Iterate.v}

We can prove that this definition satisfies \Cref{eq:iter} by defining
the function \var{rep} for repeated function application.
\lstinputlisting[
  widthgobble=2,
  firstline=8,
  lastline=12
]{../theories/BWT/Lib/Repeat.v}
In proofs, we will still use the notation $f^i z$ for $\var{rep}~f~i~z$.

@@

\var{bwl} produces the transformed string by taking the last column
of the sorted rotation matrix. \var{lexsort} sorts the rotation
matrix lexicographically using the \var{Ord} instance; we will
discuss its implementation in \cref{sec:sort}.
\lstinputlisting[
  widthgobble=2,
  firstline=31,
  lastline=35
]{../theories/BWT/BurrowsWheeler.v}

@@

\section{The inverse BWT}
\label{sec:inverse_BWT}

@@

We also can notice that the columns to the right of the sorted column
move with it for similar reasons. The column directly to its right is
the result of sorting the input string, and then stably sorting it
again by the character that precedes it in the original string. The
next column is the result of sorting by each character, sorting by the
preceding character, and then sorting by the character 2 spaces in
front.

All this means that the columns to the left of the sorted column seem
like they don't matter, which suggests that we can recreate the entire
rotation matrix from just the last column (\Cref{fig:recreate}).
However, the column directly to the left of the sorted column is very
important, as it is the next column to be sorted. And we can see that
this column changes, only becoming the eventual last column of the
matrix in the second to last iteration. Why then, does it still work
if we use the last column everywhere?

We can think of the process of sorting the matrix as computing the
last column subject to the constraints of the lexicographic ordering.
For example, we know that the collection of all characters preceding
an \texttt{a} in the sorted column must be some permutation of
$[\texttt{c},~\texttt{d},~\texttt{r},~\texttt{r},~\texttt{!}]$,
because those are the characters that precede \texttt{a}s in the
original string. However, the order of these characters in the final
string
($[\texttt{r},~\texttt{d},~\texttt{!},~\texttt{r},~\texttt{c}]$) is
important. An \texttt{r} must be first, because in the original string
there is an \texttt{a} following an \texttt{r} which is itself
followed by a \texttt{!}, the smallest character. Similarly, although
the \texttt{d} and \texttt{!} precede \texttt{a}s that are both
followed by \texttt{bra}s, the next character for the \texttt{d} is a
\texttt{!}, and the next character for the \texttt{!} is a \texttt{c}.
This requires that the \texttt{d} precede the \texttt{!} in the last
column.

@@

The word \textit{permutation} has two related meanings that are often
used interchangeably. We sometimes say that two lists are permutations of each
other if they contain all of the same elements, just appearing in a
different order. However, in mathematics we often say that a
permutation is actually the bijection on indices that defines the
correspondence between the two lists. For example, we say that $p$ is
the permutation that relates $l$ to $l'$ if for all indices $i$ and
$j$,
\begin{equation*}
  l_{p(i)} = l'_i
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
each other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
permutations of each other is precisely when
\begin{itemize}
  \item $\var{length}~l = \var{length}~l' = n$
  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
    \begin{equation*}
      l_{p(i)} = l'_i
    \end{equation*}
\end{itemize}

@@

To restrict the range to $\mathbb{N}_{<n}$ we say that $\forall i,~i \in p
\implies ~i<n$; to make the function surjective, we strengthen that to
$\forall i,~i \in p \iff ~i<n$; and to make it injective we require that $p$ not
contain any duplicate entries. The inductive property \var{NoDup} from
the Coq standard library captures that requirement:
\begin{align*}
  \infer[\var{NoDup\_nil}]{\var{NoDup}~[]}{}
  &&
  \infer[\var{NoDup\_cons}]{\var{NoDup}~(x :: l)}{%
    x \notin l & \var{NoDup}~l
  }
\end{align*}

@@

\begin{theorem}[apply\_correct]
  For all permutation functions $p$ and lists $l$,
  \begin{equation*}
    \var{Permutation}~(\var{apply}~p~l)~l
  \end{equation*}
\end{theorem}
\begin{proof}
  First we let $n = \var{length}~l$, so that we can reason by
  induction on $n$. The base case is uninteresting. In the inductive
  case, we have the induction hypothesis that for $p :
  \var{list}~\mathbb{N}$ and $l : \var{list}~A$
  \begin{equation*}
    \var{lenth}~l = n \implies \var{PermFun}~n~p \implies
    \var{Permutation}~(\var{apply}~p~l)~l
  \end{equation*}
  Given $\var{PermFun}~(n+1)~(i::p')$, we must show that
  \begin{equation*}
    \var{Permutation}~(\var{apply}~(i :: p')~(a :: l))~(a :: l)
  \end{equation*}

  We reason as follows, using the notation $\eqperm$ to
  chain together \var{Permutation}s:
  \begin{align*}
    &\var{apply}~(i :: p')~(a :: l) \\
    \eqperm{}& \text{\{ by \var{rem\_PermFun\_correct} \}} \\
    &(a :: l)_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a::l))
    \\
    \eqperm{}& \text{\{ by IH \}} \\
    &(a :: l)_i :: \var{rem\_nth}~i~(a::l)
    \\
    \eqperm{}& \text{\{ \var{rem\_nth\_Perm} \}} \\
    & a :: l \qedhere
  \end{align*}
\end{proof}

@@

For completeness, we will briefly mention a third definition of a
permutation relation that says that two lists are permutations of one
another when any element occurs an equal number of times in each:
\lstinputlisting[
  widthgobble=2, firstline=14, lastline=15,
]{../theories/BWT/Sorting/PermutationCount.v}
\var{count\_occ} counts the number of occurrences of a given variable
using a decision procedure for Leibniz equality. Here that decision
procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
require the equivalence relation itself by Leibniz equality.

@@

Based on this, we define a stronger version of the \var{PermFun}
property, \var{StablePermFun}. However, while $\var{PermFun}~n~p$
means that $p$ represents a permutation function on lists of length
$n$, we need to specify the list with respect to which $p$ is stable,
so we say $\var{StablePermFun}~l~p$ for some list $l$.
\lstinputlisting[ widthgobble=2, firstline=557, lastline=561,
]{../theories/BWT/Sorting/PermFun.v} As above, we can also define this
property using the inverse permutation: \lstinputlisting[
  widthgobble=2, firstline=563, lastline=567,
]{../theories/BWT/Sorting/PermFun.v} As usual, we will prove these
definitions equivalent. First, though, we need to define \var{image}
and \var{preimage}. \lstinputlisting[ widthgobble=2, firstline=85,
  lastline=86, ]{../theories/BWT/Sorting/PermFun.v} We will not prove
it here, but $\var{preimage}~p$ and $\var{image}~p$ are both of
bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and are
inverses of each other. We also have the following two theorems:
\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
  For all lists $l$, permutations $p$, and indices $i$,
  \begin{align*}
    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
  \end{align*}
\end{theorem}
\begin{proof}
  The first equation (\var{nth\_preimage\_apply}) holds because:
  \begin{align*}
       (\var{apply}~p~l)_{p^{-1}(i)}
    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
    &= l_{p_{(\var{findIndex}~i~p)}} \\
    &= l_i
  \end{align*}
  and the second follows from this:
  \begin{equation*}
      (\var{apply}~p~l)_i
    = (\var{apply}~p~l)_{p^{-1}(p(i))}
    = l_{p(i)}
    \qedhere
  \end{equation*}
\end{proof}

@@

Unfortunately, neither of these two definitions will be particularly
useful for proving the theorems we would like to prove with them. The
best definition for stability does not come by analogy with
\var{Permutation}. The intuition is that for any two lists that are
stable permutations of each other, if we filter each list by
equivalence with any element, the resulting lists should be identical.
This is because they must have the same elements, since the original
lists are permutations of one another, and those elements must be in
the same order, since the original lists are specifically stable
permutations of one another.
\lstinputlisting[
  widthgobble=2, firstline=21, lastline=22,
]{../theories/BWT/Sorting/StablePerm.v}

@@

\section{Uniqueness of stable sorts}
\label{sec:unique}

@@

  First, we can say that $h \equiv h'$: we know that $h \in h' :: t'$ and $h'
  \in h :: t$. If $h = h'$, then we are done, so assuming $h \neq h'$ we
  know $h \in t'$ and $h' \in t$. Because both $h::t$ and $h'::t'$ are
  sorted, we know that $h \le h'$ and $h' \le h$, implying $h \equiv h'$.

@@

\section{Lexicographic orders and keys}
\label{sec:lex_ord}

@@

First we want to prove that \var{radixsort} permutes the rows of the
matrix. We will start by proving the invariant that if we rotate the
last column to the front and sort $j$ times, the resulting matrix is a
permutation of simply rotating the last column to the front $j$
times.
\begin{lemma}[radixsort\_perm\_inv]
  For all matrices $m$ and $j : \mathbb{N}$,
  \begin{equation*}
    \var{Permutation}~((\var{map}~\var{rrot})^j~m)~
                      ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
  \end{equation*}
\end{lemma}
\begin{proof}
  We reason by induction on $j$. In the inductive case,
  \begin{align*}
    (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m ={}&
    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
    \var{map}~\var{rrot})^j~m) \\
    \eqperm{}& \var{map}~\var{rrot}~((\var{hdsort} \comp
    \var{map}~\var{rrot})^j~m) \\
    \eqperm{}& \var{map}~\var{rrot}~((\var{map}~\var{rrot})^j~m) \\
    ={}& (\var{map}~\var{rrot})^{j+1}~m \qedhere
  \end{align*}
\end{proof}

@@

\section{Relationship to the standard BWT}
\label{sec:opt}

It is important to note that standard implementations of the BWT do
not actually reconstruct the entire rotation matrix, as this would be
inefficient. Instead, they avoid this using the optimization described
in this section. Although, we did not prove or implement this
optimization in Coq, it would be straightforward to do so.

In \var{recreate}, we repeatedly prepend the last column to the front
of the matrix and sort on it. However, each time that column is the
same, so the permutation that takes us from the unsorted column to the
sorted column is the same. Therefore, we can calculate that
permutation function once and store it in a variable $p$. The
$i$th column of the recreated matrix is simply
$(\var{apply}~p)^i~(\var{bwl}~l)$, so we can define
\begin{lstlisting}
Definition recreate' (p : list nat) (l : list A) : list (list A) :=
  transpose (iter (apply p) (length l) (apply p l))
\end{lstlisting}
In the definition above, we are repeatedly applying \var{p} to the
first column of the rotation matrix, $\var{apply}~p~l$. Then we
take the transpose to create the matrix from the columns.

Since in $\var{unbwt}~i~l$, we want the $i$th row of the matrix, we
can eliminate the transpose by replacing the $\var{nth}~i$ with a
$\var{map}~\var{nth}~i$ Assuming we already have calculated $p$,
\begin{align*}
   \var{unbwt}~i~l &= (\var{recreate}~p~l)_i \\
  &= (\var{transpose}~(\var{iter}~(\var{apply}~p)~(\var{length}~l)~(\var{apply}~p~l)))_i \\
  &= \var{map}~(\lambda
  r.~r_i)~(\var{iter}~(\var{apply}~p)~(\var{length}~l)~(\var{apply}~p~l)) \\
  &= \var{map}~(\lambda r.~r_i)~(\var{tl}~(\var{iter}~(\var{apply}~p)~(1 +
  \var{length}~l)~l)) \\
  &= \var{tl}~(\var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~(1 + \var{length}~l)~l))
\end{align*}

Now, instead of indexing into each column at $i$, where the columns
are all just the result of successively applying $p$ to $l$, we will
successively take the image of $i$ under $p$ to generate the indices
into $l$ where those characters appear. This is justified via the
following theorem:
\begin{theorem}
  For all $n$,
  \begin{equation*}
  \var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~n~l)
  =
  \var{map}~(\lambda j.~l_j)~(\var{iter}~(\var{image}~p)~n~i)
  \end{equation*}
\end{theorem}
\begin{proof}
  \begin{align*}
       & \var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~n~l) \\
    ={}& \var{map}~(\lambda r.~r_i)~[l,~(\var{apply}~p)~l,~(\var{apply}~p)^2~l,~\ldots,~(\var{apply}~p)^{n-1}~l] \\
    ={}& [l_i,~((\var{apply}~p)~l)_i,~((\var{apply}~p)^2~l)_i,~\ldots,~((\var{apply}~p)^{n-1}~l)_i] \\
    ={}& [l_i,~l_{p(i)},~l_{p^2(i)},~\ldots,~l_{p^{n-1}~i}] \\
    ={}& \var{map}~(\lambda j.~l_j)~[i,~p(i),~p^2(i),~\ldots,~p^{n-1}~i] \\
    ={}& \var{map}~(\lambda j.~l_j)~(\var{iter}~(\var{image}~p)~n~i) \qedhere
  \end{align*}
\end{proof}

So assuming we have some function $\var{calc\_sort\_perm} :
\var{list}~A \to \var{list}~\mathbb{N}$ to calculate the permutation
function that sorts $l$, we redefine \var{unbwt} as
\begin{lstlisting}
Definition unbwt' (i : nat) (l : list A) : list A :=
  match l with
  | [] => []
  | d :: _ =>
    let p := calc_sort_perm l in
    let indices := tl (iter (image p) (S (length l)) i) in
    map (fun j => nth l j d) indices
\end{lstlisting}
\var{calc\_sort\_perm} can be implemented in linear time using a
counting sort, so if \var{nth} took constant time, then \var{unbwt}
would take linear time as well. As a functional specification, this
closely matches the standard imperative algorithm for the inverse BWT.

\section{Conclusions}
We have implemented and proved that the the inverse BWT actually
inverts the forwards BWT, basing our work on Bird et al.'s program
calculation derivation \cite{birdmu,pearls}. In the process, we
defined the properties of stable sorts, and implemented and proved
correct an LSD radixsort.

This paper shows that Bird's style of program calculation is very
useful for building machine-checked proofs. It leads to proofs that
are composed of many smaller lemmas, each justifying a single rewrite
step. Most of the time and effort went into proving the correctness of
radixsort. If we exclude the parts related to sorting or to other
general-purpose functions not specific to the BWT, the entire verified
implementation of the BWT consists of 42 lines of Gallina definitions,
7 lines of Gallina that are used only in proofs, and 823 lines of
theorem statements and Ltac code.\footnote{This includes
  \var{BurrowsWheeler.v}, \var{Columns.v}, \var{Rotations/Rotation.v},
  \var{Rotations/Rots.v}, \var{Lib/Iterate.v}, \var{Lib/Repeat.v} and
  \var{Lib/FindIndex.v}, which implement \var{bwl}, \var{bwn},
  \var{unbwt}, \var{cols}, \var{prepend\_col}, \var{lrot}, \var{rrot},
  \var{rots}, \var{iter}, and \var{rep}, \var{findIndex}.}

This implementation, including the optimizations mentioned in
\cref{sec:opt}, can be used as a functional specification to for a
verified implementation of bzip2. Of course, it would also be
necessary to have functional specifications for the other phases of
bzip2, such as Run-Length Encoding, Move-to-Front Encoding, and
arithmetic coding. As luck would have it, Bird has already analyzed
arithmetic coding \cite{pearls}.
\clearpage

@@

\section{Insertion sort}
\label{appendix:insertion_sort}

Insertion sort is a simple and elegant stable sort that is easy to
reason about. The theorems proved in \cref{sec:unique} allow us to
prove things using insertion sort that apply to any stable sort. For
example, our proof of correctness for radixsort relies on the fact
that it uses insertion sort as the inner sort, but
\var{StableSort\_unique} allows us to extend that proof to a radixsort
that uses any stable sort. We also use it as the implementation of
\var{lexsort} because it is convenient, but similarly
\var{Sort\_Ord\_unique} ensures that it can be replaced with any other
sort.

@@

\section{Permutation composition}
\label{appendix:perm_comp}

@@

We will use the function \var{combine} defined in the Coq standard
library to annotate every element of $l$ with its index.
\begin{lstlisting}
  Fixpoint combine (l : list A) (l' : list B) : list (A*B) :=
    match l,l' with
    | x :: tl, y :: tl' => (x,y) :: (combine tl tl')
    | _, _ => nil
    end.
\end{lstlisting}
